Northern Pikeminnow Predation on Juvenile Chinook Salmon and Potential Impacts to Local Populations in the Upper Salmon River, Idaho

### 2021-12-21 analysis/paper/paper.Rmd
incorporated edits from Google Doc

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and fall 2020 sampling periods. However, to estimate the abundance of Northern Pikeminnow during spring 2021, which was not a mark-recapture design, we used the ratio of total CPUE in the fall sampling events to the total CPUE in the spring sampling event, then multiplied that ratio by the average abundance from the fall sampling events.<!-- Kevin, is this previous statement accurate? --> This approach assumes equal capture probabilities between the fall and spring capture events. <!--If that assumption is not true and capture probability did differ, we believe that the capture probability  in the spring would be lower due to higher flows, which would lead to our spring abundance estimate to be biased low. Therefore, we presume our abundance estimate of Northern Pikeminnow for the spring 2021 sampling event to be a conservative estimate.-->

@@

<!--
Juvenile Chinook Salmon abundances were estimated at 5 rotary screw traps in the upper Salmon River basin: Sawtooth, Pahsimeroi, Hayden Creek, Upper Lemhi, and Lower Lemhi. Abundances were calculated by Idaho Department of Fish and Game at each trap which made up the production estimate for juvenile Chinook Salmon in the Upper Salmon Basin (Table X) (McClure et al. 2020).
-->

@@

@@

The PSD for Northern Pikeminnow in our study and across all three years was `r prettyNum(psd, digits = 1)`%. In other words, `r prettyNum(psd, digits = 1)`% of captured fish were over the quality size classification of `r prettyNum(q_size, digits = 0)` mm TL, demonstrating that a high percentage of Northern Pikeminnow within Deadwater Slough are above average size classes defined for the species

@@

We examined the stomach contents of `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow over the course of the study using gastric lavage methods. Of those, we documented some form of stomach contents in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` of fish of which we were able to confirm fish or fish remnants in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")`. Expressed as a percentage, `r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`% of lavaged Northern Pikeminnow had stomach contents of which `r prettyNum((gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`% we identified containing fish or fish remnants. Table \@ref(tab:gastric-tab) provides a further breakdown of gastric lavage results for each sampling event.

For Northern Pikeminnow with some form of stomach contents, the average wet weight of total contents was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; range = `r prettyNum(min(stmch_df$stomach_contents_weight), digits = 3)`-`r prettyNum(max(stmch_df$stomach_contents_weight), digits = 3)` g). Overall, fish or fish remnants were `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall weighted weight among all stomach contents examined.

@@

We estimated a single average-sized Northern Pikeminnow to consume 61 g of ***fish*** during the modeled fall time period (September 1 - November 17). Given an estimated population size of ***xx,xxx***, we'd estimate that ***xxx,xxx*** g of fish are consumed each fall during the emigration of juvenile DSR Chinook Salmon. The average weight of a DSR Chinook Salmon is ***xx.x*** taken from ***yyy*** rotary screw traps

If we further assume an average weight of ***xx g*** for DSR Chinook Salmon

Over the fall time period a single average-sized Northern Pikeminnow is modeled to eat 61 g of fish. For the estimated population of ***xx,xxx***, then ***xxx,xxx*** g of fish are being eaten each fall during the migration of juvenile Chinook Salmon. Therefore, Northern Pikeminnow in the Deadwater Slough can potentially consume ***xx,xxx*** juvenile Chinook Salmon as a discrete population.

Over an entire year, we estimate that a single Northern Pikeminnow will eat 283 g of fish and 613 g of other food types.

<!--
* Water Temperature measurements.
    *	Hobo tidbit placed at site. 
    *	Calculated daily average water temperature for 365 days of the year.
*	Water temperature plot for the 365 average.
*	Bioenergetics results. Total mass consumed over a 365 day period. Plot of daily consumption with day of year on x and mass on y.
-->

@@

@@

@@

```{r}
# pull in SAR numbers
sar = read_csv(here("analysis/data/raw_data", 
                    "WSSCh_sar_4See.csv")) %>%
  summarise(across(c(medsar, sdsar),
                   median)) %>%
  summarise(across(c(medsar, sdsar),
               ~ . / 100))

@@

```{r gastric-tab}
gastric_tbl %>%
  kable(booktabs = T,
        format.args = list(big.mark = ","),
        caption = "Summary of gastric lavage results including the total number of fish captured and gastric lavaged. For Northern Pikeminnow that were gastric lavaged, we show the number containing stomach contents, and of those, the number containing fish contents.") %>%
  kable_styling()

@@

```{r fall-bio-fig, fig.cap = "First bioenergetics plot."}
fall_bio_p = fall_only_no_growth %>%
  ggplot() +
  aes(x = date,
      y = cum_cons_fish_g,
      colour = temperature_c) +
  geom_line(size = 2.5) +
  scale_color_distiller(palette = "Spectral",
                        direction = -1) +
  scale_x_date(breaks = "2 week") +
  scale_y_continuous(name = "Scumulative fish eaten (g)",
                     breaks = seq(0, 200, by = 10),
                     sec.axis = sec_axis(~. / 12, 
                                         name = "Potential number of Chinook Salmon smolts",
                                         breaks = seq(0, 20, by = 1))) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(color = "Temperature (C)")

@@

```{r full-yr-ng-fig, fig.cap = "Next bioenergetics plot."}
full_yr_ng_p = full_yr_no_growth %>%
  ggplot() +
  aes(x = date,
      y = cum_cons_fish_g,
      colour = temperature_c) +
  geom_line(size = 2.5) +
  scale_color_distiller(palette = "Spectral",
                        direction = -1) +
  #scale_x_date(breaks = "2 week") +
  scale_y_continuous(name = "Cumulative fish eaten (g)",
                     breaks = seq(0, 600, by = 50),
                     sec.axis = sec_axis(~. / 12,
                                         name = "Potential number of Chinook Salmon smolts",
                                         breaks = seq(0, 30, by = 2))) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(color = "Temperature (C)",
       x = "Month")

@@

```{r adult-imp-p, fig.cap = "The adult impacts plot."}
adult_impact %>%
  mutate(stage = "Smolt") %>%
  select(event_name, model,
         stage,
         starts_with("smolt")) %>%
  rename_with(.fn = ~ str_remove(., "smolt_"),
              .cols = starts_with("smolt")) %>%
  bind_rows(adult_impact %>%
              mutate(stage = "Adult Equivalent") %>%
              select(event_name, model,
                     stage,
                     starts_with("adult")) %>%
              rename_with(.fn = ~ str_remove(., "adult_"),
                          .cols = starts_with("adult"))) %>%
  filter(str_detect(model, "Schuma", negate = T)) %>%
  ggplot(aes(x = model,
             y = N,
             color = model)) +
  geom_errorbar(aes(ymin = Lci,
                    ymax = Uci),
                width = 0.5) +
  geom_point(size = 4) +
  facet_grid(stage ~ event_name,
             scales = "free_y") +
  scale_color_brewer(palette = "Set1",
                     name = "Model") +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom") +
  labs(x = "N. Pikeminnow Abundance Model")
```

>>>>

The mark-recapture estimators allowed us to estimate the abundance of Northern Pikeminnow during the fall 2019 and fall 2020 sampling periods. However, to estimate the abundance of Northern Pikeminnow during spring 2021, which was not a mark-recapture design, we used the ratio of total CPUE in the fall sampling events to the total CPUE in the spring sampling event, then multiplied that ratio by the average abundance from the fall sampling events.<!-- Kevin, is this previous statement accurate? --> This approach assumes equal capture probabilities between the fall and spring capture events.

@@

@@

<!--
## Juvenile Chinook Salmon Abundance

Juvenile Chinook Salmon abundances were estimated at 5 rotary screw traps in the upper Salmon River basin: Sawtooth, Pahsimeroi, Hayden Creek, Upper Lemhi, and Lower Lemhi. Abundances were calculated by Idaho Department of Fish and Game at each trap which made up the production estimate for juvenile Chinook Salmon in the Upper Salmon Basin (Table X) (McClure et al. 2020).
-->

@@

@@

We examined the stomach contents of `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow over the course of the study using gastric lavage methods. Of those, we documented some form of stomach contents in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` of the fish and we were able to confirm fish or fish remnants in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")`. Expressed as a percentage, `r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`% of lavaged Northern Pikeminnow had stomach contents of which `r prettyNum((gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`% we identified containing fish or fish remnants. Table \@ref(tab:gastric-tab) provides a further breakdown of gastric lavage results for each sampling event.

For Northern Pikeminnow with some form of stomach contents, the average wet weight of total contents was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; range = `r prettyNum(min(stmch_df$stomach_contents_weight), digits = 3)`-`r prettyNum(max(stmch_df$stomach_contents_weight), digits = 3)` g). Overall, fish or fish remnants were `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of the overall weight among all stomach contents examined.

@@

We estimated a single average-sized Northern Pikeminnow to consume xx.x g of fish during the modeled fall time period (September 15 - November 30) and an additional xx.x g of fish during the modeled spring time period (March 1 - June 31). Given estimated population sizes of xx,xxx Northern Pikeminnow during fall and xx,xxx during spring, that would convert to xxx,xxx - xxx,xxx g during fall and xxx,xxx g during spring. Given average weights of 10.3 g for DSR Chinook salmon and 10.9 g for NRR Chinook salmon, we would estimate that xx,xxx and xx,xxx juvenile Chinook salmon could be consumed during the fall and spring emigrations in Deadwater Slough, respectively.

<!--
For context, throughout the migration periods of DSR and NRR life histories it is estimated that 113,976 individuals are produced and therefore available within the upper Salmon River basin (Table x).

Given an estimated population size of xx,xxx, we’d estimate that xxx,xxx g of NRR fish are consumed each spring during the emigration, and xxxxx.xx g of DSR fish are consumed in the fall during the emigration of juvenile Chinook Salmon. Therefore, Northern Pikeminnow in the Deadwater Slough can potentially consume xx,xxx juvenile Chinook Salmon as a discrete population.
-->

<!--
## Juvenile Chinook Salmon Abundance

The abundance of juvenile Chinook Salmon for the upper Salmon River basin was estimated at 6 rotary screw traps [@McClure2020]. Overall, in 2020, there were an estimated 59,289 DSR, 22,264 NRR, and 10,159 parr estimated among all these traps in the upper Salmon River basin (Table X.)
-->

@@

<!--
The PSD for Northern Pikeminnow in our study and across all three years was `r prettyNum(psd, digits = 1)`% i.e., `r prettyNum(psd, digits = 1)` of captured fish were over the quality size classification of `r prettyNum(q_size, digits = 0)` mm TL defined by @Winther2020. For comparison, @Winther2020 reported a PSD for Northern Pikeminnow captured below Bonneville Dam of 41% and 18% in Bonneville Reservoir, suggesting that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to population reported elsewhere in the Columbia River. 
-->

@@

<!--If that assumption is not true and capture probability did differ, we believe that the capture probability  in the spring would be lower due to higher flows, which would lead to our spring abundance estimate to be biased low. Therefore, we presume our abundance estimate of Northern Pikeminnow for the spring 2021 sampling event to be a conservative estimate.-->

@@

```{r, eval = F}
# pull in SAR numbers
sar = read_csv(here("analysis/data/raw_data", 
                    "WSSCh_sar_4See.csv")) %>%
  summarise(across(c(medsar, sdsar),
                   median)) %>%
  summarise(across(c(medsar, sdsar),
               ~ . / 100))

@@

```{r gastric-tab}
gastric_tbl %>%
  rename(Effort = event_name,
         Captured = captured,
         Lavaged = lavaged,
         'Stomach Contents' = stmch_contents,
         'Fish Contents' = fish_contents) %>%
  mutate(Effort = recode(Effort,
                         `Fall_2019` = "Fall 2019",
                         `Fall_2020` = "Fall 2020",
                         `Spring_2021` = "Spring 2021")) %>%
  kable(booktabs = T,
        format.args = list(big.mark = ","),
        caption = "Summary of gastric lavage results including the total number of fish captured and gastric lavaged. For Northern Pikeminnow that were gastric lavaged, we show the number containing stomach contents, and of those, the number containing fish contents.") %>%
  kable_styling()

@@

```{r fall-bio-fig, eval = F, fig.cap = "First bioenergetics plot."}
fall_bio_p = fall_only_no_growth %>%
  ggplot() +
  aes(x = date,
      y = cum_cons_fish_g,
      colour = temperature_c) +
  geom_line(size = 2.5) +
  scale_color_distiller(palette = "Spectral",
                        direction = -1) +
  scale_x_date(breaks = "2 week") +
  scale_y_continuous(name = "Scumulative fish eaten (g)",
                     breaks = seq(0, 200, by = 10),
                     sec.axis = sec_axis(~. / 12, 
                                         name = "Potential number of Chinook Salmon smolts",
                                         breaks = seq(0, 20, by = 1))) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(color = "Temperature (C)")

@@

```{r full-yr-ng-fig, eval = F, fig.cap = "Next bioenergetics plot."}
full_yr_ng_p = full_yr_no_growth %>%
  ggplot() +
  aes(x = date,
      y = cum_cons_fish_g,
      colour = temperature_c) +
  geom_line(size = 2.5) +
  scale_color_distiller(palette = "Spectral",
                        direction = -1) +
  #scale_x_date(breaks = "2 week") +
  scale_y_continuous(name = "Cumulative fish eaten (g)",
                     breaks = seq(0, 600, by = 50),
                     sec.axis = sec_axis(~. / 12,
                                         name = "Potential number of Chinook Salmon smolts",
                                         breaks = seq(0, 30, by = 2))) +
  theme_classic() +
  theme(legend.position = "top") +
  labs(color = "Temperature (C)",
       x = "Month")

@@

```{r adult-imp-p, eval = F, fig.cap = "The adult impacts plot."}
adult_impact %>%
  mutate(stage = "Smolt") %>%
  select(event_name, model,
         stage,
         starts_with("smolt")) %>%
  rename_with(.fn = ~ str_remove(., "smolt_"),
              .cols = starts_with("smolt")) %>%
  bind_rows(adult_impact %>%
              mutate(stage = "Adult Equivalent") %>%
              select(event_name, model,
                     stage,
                     starts_with("adult")) %>%
              rename_with(.fn = ~ str_remove(., "adult_"),
                          .cols = starts_with("adult"))) %>%
  filter(str_detect(model, "Schuma", negate = T)) %>%
  ggplot(aes(x = model,
             y = N,
             color = model)) +
  geom_errorbar(aes(ymin = Lci,
                    ymax = Uci),
                width = 0.5) +
  geom_point(size = 4) +
  facet_grid(stage ~ event_name,
             scales = "free_y") +
  scale_color_brewer(palette = "Set1",
                     name = "Model") +
  theme(axis.text.x = element_blank(),
        legend.position = "bottom") +
  labs(x = "N. Pikeminnow Abundance Model")
```
